|

### 2021-07-22 prova_knit_finale.Rmd
auxiliry fixed

\begin{algorithm} Kalman Filter for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute the one-step-ahead state predictive distribution at time $t-1$, notice that if $t=1$ no data have been observed yet and therefore $x_{1}|x_{0}\sim N(a_{1},R_{1})$ otherwise
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e. the product of the one-step-ahead state predictive distribution and the likelihood
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{DLM} function implement in R replicate this steps.
\
\hrule
\hrule
\
\code{DLM(data,sig2,tau2,m0,C0)}
\
\hrule
\textbf{Arguments}

\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
\
\code{sig2} \ \ the variance $\sigma^{2}$ in observation equation
\
\code{tau2} \ \ the variance $\tau^{2}$ in state equation
\
\code{m0} \ \ central value of the normal prior state distribution
\
\code{C0} \ \ variance of the normal prior state distribution
\hrule
\hrule
```{r}
DLM<-function(data,sig2,tau2,m0,C0){
  n  = length(data)
  m  = rep(0,n)
  C  = rep(0,n)
  for (t in 1:n){
    if (t==1){
      a = m0
      R = C0 + tau2
    }else{
      a = m[t-1]
      R = C[t-1] + tau2
    }
    A = R/(R+sig2)
    m[t] = (1-A)*a + A*y[t]
    C[t] = A*sig2
  }
  return(list(m=m,C=C))
}
```
In Figure \ref{fig:myfig2} below filtered states estimated using Kalman Filter with $x_{0} \sim N(0,100)$ and $\sigma^{2}=\tau^{2}=1$ are compared to the true states values.
Notice how closely the filtered states follow the observations and the goodness of the approximation of the true states. 95 percent credible intervals are computed as
\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
```{r myfig2, echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

@@

```{r echo=FALSE, fig.align='center', fig.cap="a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
FILT.GPF<-Filtplot(df,gpf,"")
ggarrange(FILT.GPF,ESS.GPF,labels=c("(a)","(b)"))
```
Let's provide directly a brief comparison between the Bootstrap Particle Filter (BPF) and the Guided Particle Filter (GPF). As we can see from the Figure \ref{fig:myfig5}, the Guided Particle Filter provides point estimates that are slightly better with respect to the ones of the BPF, and this is confirmed by the Table showing the RMSE. Moreover, also the variance is better, suggesting for higher precision.

@@

    if(ESS<(N/r)){
    weight = w*dnorm(data[t],x,predsd)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    }else{
    weight = rep(1/N,N)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    }

    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
    x1   = rnorm(N,means,importancesd)
    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)
    w   = exp(lw)
    w   = w/sum(w)
    x <- x1

@@

    weight = w*dnorm(data[t],x,predsd)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
    x1   = rnorm(N,means,importancesd)
    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)

@@

@@

We decide to organise this Appendix as an R package. All the R functions described in this Appendix take as reference a basic specification of Stochastic Volatility Model.

\begin{align*}
    y_t|x_t & \sim N(0,e^{x_t})\\
    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align*}

All the plots and tables of this work concerning stochastic volatility models use this functions.

@@

>>>>

\begin{algorithm} Kalman Filter for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $x_{0} \sim N(m_{0},C_{0})$
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute the one-step-ahead state predictive distribution at time $t-1$, notice that if $t=1$ no data have been observed yet and therefore $x_{1}|x_{0}\sim N(a_{1},R_{1})$ otherwise
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e. the product of the one-step-ahead state predictive distribution and the likelihood
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{DLM} function implement in R replicate this steps.
\
\hrule
\hrule
\
\code{DLM(data,sig2,tau2,m0,C0)}
\
\hrule
\textbf{Arguments}

\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
\
\code{sig2} \ \ the variance $\sigma^{2}$ in observation equation
\
\code{tau2} \ \ the variance $\tau^{2}$ in state equation
\
\code{m0} \ \ central value of the normal prior state distribution
\
\code{C0} \ \ variance of the normal prior state distribution
\hrule
\hrule
```{r}
DLM<-function(data,sig2,tau2,m0,C0){
  n  = length(data)
  m  = rep(0,n)
  C  = rep(0,n)
  for (t in 1:n){
    if (t==1){
      a = m0
      R = C0 + tau2
    }else{
      a = m[t-1]
      R = C[t-1] + tau2
    }
    A = R/(R+sig2)
    m[t] = (1-A)*a + A*y[t]
    C[t] = A*sig2
  }
  return(list(m=m,C=C))
}
```
In Figure \ref{fig:myfig2} below filtered states estimated using Kalman Filter with $x_{0} \sim N(0,100)$ and $\sigma^{2}=\tau^{2}=1$ are compared to the true states values.
Notice how closely the filtered states follow the observations and the goodness of the approximation of the true states. 95 percent credible intervals are computed as
\[[E(x_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(x_{t}|y_{1:t})},E(x_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(x_{t}|y_{1:t})}]\]
```{r myfig2, echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

@@

```{r echo=FALSE, fig.align='center', fig.cap="a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
FILT.GPF<-Filtplot(df,gpf,"")
ggarrange(FILT.GPF,ESS.GPF,labels=c("(a)","(b)"))
```
Let's provide directly a brief comparison between the Bootstrap Particle Filter (BPF) and the Guided Particle Filter (GPF). As we can see from the Figure \ref{fig:myfig5}, the Guided Particle Filter provides point estimates that are slightly better with respect to the ones of the BPF, and this is confirmed by the Table showing the RMSE. Moreover, the variance is lower, suggesting for higher precision.

@@

    if(ESS<(N/r)){
    weight = w*dnorm(data[t],x,predsd)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
    x1   = rnorm(N,means,importancesd)
    w   = rep(1/N,N)
    }else{
    weight = rep(1/N,N)
    k   = sample(1:N,size=N,replace=FALSE,prob=weight)
    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
    x1   = rnorm(N,means,importancesd)
    w   <-  w*dnorm(data[t],x1,predsd)
    }

    w   = w/sum(w)
    x <- x1

@@

@@

So far we have seen Sequential Monte Carlo methods applied to the linear Gaussian case, however, as already mentioned, for this class of state space models, Kalman Filter provides a closed form solution

@@

We decide to organise this Appendix as an R package. All the R functions described in this Appendix take as reference a basic specification of Stochastic Volatility Model. 
\begin{align*}
    y_t|x_t & \sim N(0,e^{x_t})\\
    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
\end{align*}
All the plots and tables of this work concerning stochastic volatility models use this functions.

@@

    weight = w*dnorm(data[t],x,predsd)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
    x1   = rnorm(N,means,importancesd)
    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)+log(w)-log(weight)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)
